---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# crosstabser

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/crosstabser)](https://CRAN.R-project.org/package=crosstabser)
[![crosstabser status badge](https://urswilke.r-universe.dev/crosstabser/badges/version)](https://urswilke.r-universe.dev/crosstabser)
<!-- badges: end -->

The goal of crosstabser is to generate crosstabs from labelled data (usually survey data).
Its parent dependency [datadaptor](https://gitlab.com/urswilke/datadaptor) 
uses a similar approach to modify datasets. 
The generated crosstabs can be plotted interactively in html files with the apps
provided by [table_charter](https://gitlab.com/urswilke/table_charter).
<!-- TODO: refer to table books on our website!-->
<!-- TODO: refer to spreadator!-->


## Installation

You can install the development version of crosstabser like so:

```{r eval=FALSE}
devtools::install_gitlab("urswilke/crosstabser")
```

## Example

First we'll load the package.

```{r example}
library(crosstabser)
```


### Minimal example

Suppose you have survey data `df`.

```{r}
df <- tibble::tibble(
  q1 = c(1, 2, 1) |> haven::labelled(c(Yes = 1, No = 2), label = "Super important question"),
  age = c(2, 1, 1) |> haven::labelled(c("18-39" = 1, "40+" = 2), label = "age")
)
df
```

The data consists of 3 cases with the answers Yes & No stored in the variable `q1` and the age category stored in the variable `age`.

And suppose you want to generate a crosstab, summarizing the values of `q1` on the y-axis, and on the x-axis showing the total counts, as well as those in the sub-populations of the variable `age`.

crosstabser allows this if you define a mapping object like this:

```{r}
(mapping_file = list(
  Questions = data.frame(
    Type  = "cat",
    RowVar = "q1",
    Title = "The crosstab's title"
  ),
  Macro = list(ColVar = "age")
))
```

Then you can use the package's `Tabula` class to generate print output of the crosstab in the console:

```{r}
Tabula$new(df, mapping_file)
```

Please refer to `vignette("questions")` for some more examples, 
and `vignette("questions-parameters")` 
for details how to use the parameters in the mapping.
[Here](https://urswilke.github.io/datadaptor-crosstabser-table_charter-demo/)
you can find an interactive introduction 
that also shows the resulting table_charter app.


