---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# crosstabser

<!-- badges: start -->
<!-- badges: end -->

The goal of crosstabser is to ...

## Installation

You can install the development version of crosstabser like so:

```{r eval=FALSE}
devtools::install_gitlab("urswilke/crosstabser")
```

## Example

First we'll load the package.

```{r example}
library(crosstabser)
```


### Minimal example

Suppose you have survey data `df`.
```{r}
df <- tibble::tibble(
  q1 = c(1, 2, 1) |> haven::labelled(c(Yes = 1, No = 2), label = "Super important question"),
  age = c(2, 1, 1) |> haven::labelled(c("18-39" = 1, "40+" = 2), label = "age")
)
df
```
The data consists of 3 cases with the answers Yes & No stored in the variable `q1` and the age category stored in the variable `age`.

And suppose you want to generate a crosstab, summarizing the values of `q1` on the y-axis, and on the x-axis showing the total counts, as well as those in the sub-populations of the variable `age`.

crosstabser allows this if you define a mapping object like this:
```{r}
(mapping_file = list(Questions = data.frame(
  Type  = "cat",
  RowVar = "q1",
  Title = "The crosstab's title"
)))
```
This is a list with one element `"Questions"` containing a one row data.frame.
Then you can use the package's `Tabula` class to generate print output of the crosstab in the console:
```{r}
Tabula$new(df, mapping_file, colvar = "age")
```

