---
title: "Format of the crosstabs' data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Format of the crosstabs' data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


In this vignette we'll show the underlying data format of the crosstabs generated by crosstabser.
Basically, it's a long format of 
[tidy data](https://r4ds.had.co.nz/tidy-data.html) 
such as the one needed if you'd want to plot the crosstab with
[ggplot2](https://github.com/tidyverse/ggplot2/) or 
[observable plot](https://github.com/observablehq/plot) (like in table_charter).
However, in order to reduce redundancy and to save space
we store the data in multiple data.frames that can be merged together in the end.
First, let's load the needed libraries:

```{r, setup, message=FALSE}
library(crosstabser)
library(dplyr)
library(ggplot2)
library(purrr)
library(tidyr)
library(haven)
```

and define a labelled data.frame:

```{r}
df <- tibble::tibble(
  q1 = c(1, 2, 1) |> haven::labelled(c(Yes = 1, No = 2), label = "Super important question"),
  age = c(2, 1, 1) |> haven::labelled(c("18-39" = 1, "40+" = 2), label = "age")
)
df
```


Next we define the syntax to generate a crosstab:
```{r}
Questions <- tibble::tibble(
  Abbreviation = "Q1",
  Type  = "cat",
  RowVar = "q1",
  Title = "The crosstab's title"
)
```

We'll use the `age` variable for the x-axis of the crosstab:

```{r}
ColVar <- "age"
```

Now we can construct an R6 object of class "Tabula":

```{r}
mapping_file <- list(Questions = Questions, Macro = list(ColVar = ColVar))
m <- Tabula$new(
  df,
  mapping_file,
)
```

After running the `$prepare_5_tables()` method,

```{r}
m$prepare_5_tables()
```

the resulting data is stored in the `$ditw$ct$crosstabs$data` field:

```{r}
l <- m$ditw$ct$crosstabs$data
l
```
<!-- TODO: remove unneeded columns!-->
It contains 5 data.frames:

  - `tab_table`: For every crosstab generated, this dataframe contains 1 row;
    `QuestNo` is the unique identifier from `Questions$Abbreviation`; 
    if multiple crosstabs are generated by a row, they are identified by `TabNo`.
  - `val_table`: Contains the values of the crosstabs in long format in the column `Value`. 
  - `row_table`: Contains the label information of the crosstabs' rows and 
    some information about the format in the rows.
  - `head_table`: Contains the label information of the crosstabs' headers.
  - `col_table_all`: Contains the label information of the crosstabs' columns.

Now we're ready to merge all this data into one data.frame:

```{r}
res <- l[c(
  "tab_table", 
  "val_table", 
  "row_table", 
  "col_table_all",
  "head_table"
)] |> 
  reduce(left_join)

```

<details>
  <summary>Click here to see the full data.frame</summary>

```{r}
knitr::kable(res)
```
</details>



If we look at the crosstab
```{r}
m
```

and say we wanted to generated a color-coded raster of the percent values,
we could do this like this:

```{r}
res |> 
  filter(
    # This removes the data of the "TOTAL" & "VALID CASES" rows:
    RowContent == "Detail", 
    # remove rows with absolute values:
    RowAbsPercent == "Percent"
  ) |> 
  ggplot() +
  geom_tile(aes(
    x = ColTitle2,
    y = RowTitle1,
    fill = Value
  )) + 
  facet_grid(
    ~ as_factor(ColTitle1), 
    scales = "free_x"
  ) +
  scale_x_discrete(position = "top") +
  theme_minimal() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )
```

(*This plot isn't really interesting; 
this is just to demonstrate the structure of the crosstabs' underlying data*)
