---
title: "Calculating crosstabs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating crosstabs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette we'll show some examples how crosstabs can be generated.
[Here](https://urswilke.github.io/datadaptor-crosstabser-table_charter-demo/)
you can find an interactive version of this vignette 
that will also automatically generate an interactive
[table_charter](https://gitlab.com/urswilke/table_charter) app of the crosstabs.
First we'll load the library

```{r, setup}
library(crosstabser)
library(haven)
```

and define a labelled data.frame:

<details>
  <summary>Show code how the data.frame was generated</summary>

```{r}
set.seed(123)
n <- 10
df <- tibble::tibble(
    q1_1 = sample(c(0, 1), n, TRUE) |> labelled(c(Selected = 1, Unselected = 0), label = "Pizza"),
    q1_2 = sample(c(0, 1), n, TRUE) |> labelled(c(Selected = 1, Unselected = 0), label = "Lasagne"),
    q1_3 = sample(c(0, 1), n, TRUE) |> labelled(c(Selected = 1, Unselected = 0), label = "Bruschetta"),
    q2n1 = sample(c(1, 2, 3), n, TRUE) |> labelled(c(Beer=1, Sauerkraut=2, Brezel = 3), label = "1st choice"),
    q2n2 = sample(c(1, 2, 3, rep(NA_real_, 1)), n, TRUE) |> labelled(c(Beer = 1, Sauerkraut = 2, Brezel = 3), label = "2nd choice"),
    q2n3 = sample(c(1, 2, 3, rep(NA_real_, 4)), n, TRUE) |> labelled(c(Beer = 1, Sauerkraut = 2, Brezel = 3), label = "3rd choice"),
    q2n4 = sample(c(1, 2, 3, rep(NA_real_, 8)), n, TRUE) |> labelled(c(Beer = 1, Sauerkraut = 2, Brezel = 3), label = "4th choice"),
    q3_1 = sample(c(1, 2, 3, 4, 5), n, TRUE) |> labelled(c("Not at all" = 1, "2" = 2, "3" = 3, "4" = 4, "Very much" = 5), label = "Dogs are cool"),
    q3_2 = sample(c(1, 2, 3, 4, 5), n, TRUE) |> labelled(c("Not at all" = 1, "2" = 2, "3" = 3, "4" = 4, "Very much" = 5), label = "Cats are beautiful"),
    age  = sample(c(1, 2, 3), n, TRUE) |> labelled(c("18-39" = 1, "40-59" = 2, "60+" = 3), label = "age"),
    gew = runif(n, 0.5, 3),
)
# Here, we make sure that this weight variable sums up to `n`:
df$gew <- df$gew / mean(df$gew)
```
</details>

```{r}
df
```

Next we define the syntax to generate crosstabs:
```{r}
Questions <- tibble::tibble(
  Abbreviation = c("age", "fav_it", "fav_de", "item_bat"),
  Title = c(
    "Age", "What`s your favorite Italian food?",
    "What`s your favorite German food?", "How much do you agree?"
  ),
  Type = c("cat", "mdg", "mcg", "mw"),
  RowVar = c("age", "q1_1 q1_2 q1_3", "q2n1 q2n2 q2n3 q2n4", "q3_1 q3_2"),
  MdgMissLab = c("", "No answer", "", ""),
  Freq = c("", "", "", "0")
)
Questions
```

To define the columns along the x-axis of the crosstabs, we need to define a character vector of variable names in `df`:

```{r}
ColVar <- "age"
```

The default method to generate crosstabs is by using Excel mapping files. 
You could generate a template "mapping.xlsx" for such a file by using:

```{r, eval=FALSE}
create_tabula(df, "mapping.xlsx")
```
In the generated file you could then copy `Questions` into the corresponding columns of the "Questions" sheet, 
and `ColVar` into the "ColVar" row of the "Macro" sheet. There, the column that is used is defined by the number of the named region "V_Scenario" that corresponds to the "Scenario" row (this allows to store varying parameter sets for different scenarios).

However, for easier documentation the package also allows to define mappings as lists 
consisting of 2 data.frame elements `Questions` & `Macro`:
```{r}
mapping_file <- list(Questions = Questions, Macro = list(ColVar = ColVar))
```

Now we can construct an R6 object of class "Tabula":

```{r}
m <- Tabula$new(
  df,
  mapping_file,
)
```

The `Tabula` class has a special print method that prints all the crosstabs calculated:

```{r}
m
```

If you want to weight the variables by `df$gew`, here's how:
```{r}
mapping_file$Macro$Weight <- "gew"
m <- Tabula$new(
  df,
  mapping_file,
  tabulate = FALSE,
)
# (by setting `tabulate = FALSE and the next line,
# we can change options, before calculating the crosstabs.
# this changes the Symbol for the weighted rows from "⚖" to "W" for a better print output):
m$opts$ct$l_lexikon["cTabWeighted"] <- " W"
m$calc_qtabs()
m
```

You can also generate both, unweighted & weighted rows (alternating) in the tables by further adding:
```{r}
# alternatively we could have set this parameter upfront, as above:
# mapping_file$Macro$Unwgt <- TRUE
m$opts$ct$l_macro_scenario$Unwgt <- TRUE
m$calc_qtabs()
m
```

For the various options how to modify the crosstabs, please have a look at
`vignette("question-parameters")`.
