---
title: "Parameters in the Questions sheet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameters in the Questions sheet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include=FALSE}
# for inline command with backticks:
# (cf. https://stackoverflow.com/questions/72283861/rmarkdown-backticks-inside-inline-code-inconsistent-behavior-with-usual-code-c/72284466#72284466)
metr_mac_funs <- paste0("`", crosstabser:::df_metr_mac$fun, "`") |> 
  glue::glue_collapse(", ", last = " and ")
```


This vignette lists the options how to modify the crosstabs generated with crosstabser.
The items in the bullet points correspond to the column headers of the table in the Questions sheet: 

## Mandatory arguments

The arguments `Title`, `Type` and `RowVar` need to be specified for all crosstabs:

-	`Title`: Question title that will be shown in the table; 
  A blank in single quotes (`' '`) inserts a line break
<!-- TODO: use different names for the 4 types-->
-	`Type`: Defining the type of question; there are 4 possible cases:
    -	`cat`: for single-choice questions; if multiple `RowVar`s are specified, 
      this will generate multiple crosstabs (along the y-axis) with the variable
      labels of the `RowVar`s as a sub-title
    -	`mw`: for item batteries questions, 
      e.g. "Rank each item on a scale from 1 – 5."; 
      First a table is created showing the mean values of the `RowVar` variables in its lines, the "Summary of means" table. 
      Then, a series of tables showing the frequency counts of each of the categorical variables is generated similar to the case `cat`, the "detail tables". 
    -	`mdg`: multiple dichotomy group (called MDGROUP in SPSS); multiple-choice question;
      For each answer option there is one variable; 
      the text of the choice is in the variable label of the accdording variable;
      If this answer was selected the variable usually has the value 1.
    -	`mcg` : multiple category group (called  MCGROUP in SPSS); multiple-choice question; 
      This is an alternative way to `mdg` to code multiple choice questions.
      The answer possibilities are stored in the value labels here (every variable has the same value labels).
      This means, you need at least as many variables as the maximum number of choices that were selected by the respondents.
-	`RowVar`: variable names to be tabulated (space-separated); 
  alternatively, if the string starts with the prefix `"ts: "` (without the quotes), you can use
  [tidy-select](https://dplyr.tidyverse.org/reference/dplyr_tidy_select.html) syntax afterwards

## Optional arguments

### Optional arguments for all 4 `Type`s

-	`Abbreviation`: An identifier string for internal use; 
  if duplicated or not specified, `"_row_<ROW VALUE>"` will be appended 
  (where `<ROW VALUE>` is the row number)
-	`Unguelt`: values / variable names (for `"mdg"`) representing the unvalid cases 
-	`Filter`: logical expression for the cases that should be filtered
-	`SelVar` & `SelVal`:
  calculates the table(s) conditionally for every value in `SelVal` that the Variable `SelVar` takes, the variables in RowVar have to be mentioned as many times as the number of variables in `SelVar`. Format of `SelVal`: "1 2 3 4:Name_it_different 2-97:All_others" to produce a table for all values 1 to 4 and one where 2-97 are combined and labelled "All others". 
-	`Mult`: if `TRUE` in `SelVar`/`SelVal` interviews should be counted double, if `SelVar` is `SelVal` more than once (i.e. if in `SelVar` two "Other manufacturers" mentioned, they will be seen as two answers)
-	`Weight`: variable with which the values are weighted (will overwrite the one that is set in the sheet Macro if existing)
<!-- TODO: -	`WeightLab`: will be added to the title of the table in a new line, syntax unsure -->
<!-- -	ColPl: Different headers than in Sheet Macro Row ColPl can be defined (have to have the same number of columns) -->

### Optional arguments for `Type = cat`

-	`CatRec` & `CatLab`:
    - `CatRec`: Possibility to add to the table of a categorical variable (`Type = cat OR mw`) the statistics of a recoded variable summarizing categories
    - `CatLab`: Definition of the labels of the recoded summary variable in `CatRec`
<!-- TODO: Example: Summarize a categorical variable from 1 to 5, where 1: very positive, 2:positive, 3 neutral, 4 negative and 5 very negative: Reassign 1 & 2 to 1, 3 to 2, and 4 & 5 to 3: -->
<!-- CatRec: (1 THRU 2=1) (3=2) (4 THRU 5=3)|(1 THRU 34=1) -->
<!-- CatLab: 1 '1 and 2' 2 '3' 3 '4 and 5'|1 ‘2nd summary’ -->
-	`MWRec`: values can be recoded for the calculation of the mean value with the equal syntax as in `CatRec`.
-	`UngueltMW`: values that are excluded for the mean value calculation (if empty, `Unguelt` ist taken)
-	`MetrMac`: Possibility to add one or more of the following lines to the table showing statistics of the variable:
    -	`Sx`: Mean with x decimals
    -	`Mx`: Median with x decimals
    -	`Tx`: Total with x decimals
    -	`Ex`: Standard error with x decimals
    - `Ix`: Minimum with x decimals
    - `Ax`: Maximum with x decimals
    -	`Pnnx`: `nn`th Percentile with x decimals

<!-- -	MWVar: take this variable instead of rowvar for calculating the mean -->
-	`Einzelauspraegung`: if set to 0 the rows in the table corresponding to the values of the variable are removed from the table (`Type = "cat"`)


### Optional arguments for `Type = mw`

-	`MW`: if 0 the "Summary of means" table is not calculated for `Type = "mw"`
-	`Freq`: If 0 the detail tables are not calculated for `Type = "mw"`
<!-- TODO:-	Zsfg: ?? -->
-	`MeanOverviewLabel`: label for summary of means table (normally "Summary of means", cf. lexikon `cTabMeanOV`)
-	`ZsfgMW`: Function name to use in `Type = "mw"`; possible values are `r metr_mac_funs`.
- `RepOV`: tbd


### Optional arguments for `Type = mdg`

-	`MdgVal`: The value that is counted as selected for `Type = "mdg"`, can be a range (either 0-97 or 0 THRU 97)
-	`MdgMissLab`: The Label that is written for the count of cases where none of the variables is selected (`Type = mdg`); Default in English: "No entry in the selected variables" (see lexikon `cTabNoEntry`)
-	`MdgMissValid`: if TRUE missing cases are set to valid (`Type = "mdg"`)
<!-- TODO: -	`ZsfgMFA`: SPSS Syntax for rows in mdg / mcg (normally count and share of valid answers) -->

### Other arguments tbd
<!-- TODO: discuss with Wolf and fully implement-->
-	`RvEmp`: If set to `"EXCLUDE"` Rows of values in the resulting tables with 0 counts will be removed.
<!-- TODO: -	`Categories`: New feature to be programmed for `Type = mcg` tables showing a recoded summary variable as in `CatRec`, however without double-counting individuals in the new categories of the recoded variable; the categories of the recoded variable are shown in the lines above the according original categories. -->
-	`Sort`: Possibility to resort the columns showing the categories in the table alphanumerically or by share;\
  Examples:
    -	Descending by frequency of items: `ORDER=D KEY=COUNT`
    -	Ascending: `ORDER=A`
