---
title: "Calculating crosstabs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating crosstabs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

In this vignette we'll show some examples how crosstabs can be generated.
First, a labelled data set is modified with 
[datadaptor](https://gitlab.com/urswilke/datadaptor)
to prepare the data for the crosstabs.
The latter are then calculated with the crosstabser functionalities.
[Here](https://urswilke.github.io/datadaptor-crosstabser-table_charter-demo/)
you can find an interactive version of this vignette 
that will also automatically generate an interactive
[table_charter](https://gitlab.com/urswilke/table_charter) 
app of the crosstabs in the browser.


## Getting started

First, we'll load the library.

```{r, setup}
library(crosstabser)
```

In order to use crosstabser, we need a dataset and a mapping file.
The latter defines the commands to modify the data (optional; using datadaptor functionalities), 
and those to calculate crosstabs.

We'll use the `datadaptor::fruit_survey` toy data set

```{r}
df <- datadaptor::fruit_survey
```

together with the 
[example mapping file](https://gitlab.com/urswilke/crosstabser/-/blob/master/inst/extdata/tabulate-fruits.xlsx) 
for this data set in the package. 
If you install crosstabser on your machine, this file is also on your computer:

```{r}
mapping_file <- system.file("extdata", "tabulate-fruits.xlsx", package = "crosstabser")
```

We can apply the data modifications and calculate the crosstabs in one command
by defining an R6 object of class "Tabula":

```{r, message=FALSE}
m <- crosstabser::Tabula$new(
  dat = df, 
  mapping_file = mapping_file
)
```

and then print the crosstabs defined in the mapping:

(*by passing a high value of the* `width` 
*parameter, we can make sure that all headers are printed*)
```{r, R.options = list(width = 1e4)}
m |> print(width = 1e4)
```


## Dive-in

In the following, we'll have a closer look, how this works.


### Create a mapping file

crosstabser's example file `mapping_file` to generate crosstabs was built
by editing the empty template of an Excel mapping file. 
For every data set `df`, 
this is how you can generate the empty mapping file 
(named `"mapping.xlsx"` here) 
from the template:

```{r, eval=FALSE}
create_tabula(df, "mapping.xlsx")
```


### Mapping file content {.tabset}

```{r, echo=FALSE}
sheets <-   c("Variables", "Label", "Free1", "Macro", "Questions")
```

You can also see the relevant parts in the sheets of `mapping_file` here,
in the following `r length(sheets)` data.frames:

```{r, echo=FALSE}
wb <- openxlsx2::wb_load(mapping_file)
l <- purrr::map2(
  sheets |> purrr::set_names(),
  c(TRUE, TRUE, FALSE, FALSE, TRUE),
  \(x, b) openxlsx2::read_xlsx(wb, x, col_names = b, ) |> 
    dplyr::mutate(across(-matches("(^nv$|sum_var_value)"), as.character)) |> 
    tibble::as_tibble()
)
l$Variables <- l$Variables |> 
  dplyr::filter(if_any(c("new_label", "new_name", "op"), Negate(is.na)))
l$Label <- l$Label |> 
  dplyr::filter(if_any(c("new_label", "sum_var_label", "sum_var_value", "sum_var_vallab"), Negate(is.na)))
new_names <- paste0("X", 1:5)
names(l$Free1) <- new_names
l$Free1 <- l$Free1 |> 
  dplyr::mutate(row = dplyr::row_number(), .before = 1) |> 
  dplyr::filter(if_any(all_of(new_names), Negate(is.na)))
l$Macro <- l$Macro |> janitor::remove_empty("cols") |> _[1:2] |> tidyr::drop_na(E)
l$Questions <- l$Questions |> janitor::remove_empty("cols") |> janitor::remove_empty("rows")
not_needed <- list2env(l, globalenv())
```


#### Variables

```{r, R.options = list(width = 1e4)}
Variables
```

#### Label

```{r, R.options = list(width = 1e4)}
Label
```

#### Free1

```{r, R.options = list(width = 1e4)}
Free1
```

#### Macro

```{r, R.options = list(width = 1e4)}
Macro
```

We only need the column 
where `Scenario` is equal to the value `"V_Scenario"` specified in the mapping. 

(This allows to store varying parameter sets for different scenarios. 
And by using this named region in Excel formulas, 
you can also enter scenario-specific commands in the Excel file.)

The only relevant argument in this case is the value in the `"ColVar"` row.
The variable names are entered in a space separated list, 
corresponding to this character vector:

```{r, comment='', echo=FALSE}
colvar_cmd <- l$Macro |> 
  dplyr::filter(A == "ColVar") |>
  _[-1][[1]] |> 
  strsplit(" +") |>
  _[[1]] |> 
  dput() |> capture.output()
colvar_cmd <- paste("ColVar <- ", colvar_cmd, collapse = "")
eval(parse(text = colvar_cmd))
cat(colvar_cmd)
```


#### Questions {.active}

```{r, R.options = list(width = 1e4)}
Questions
```

Only the first three columns `Title`, `RowVar` & `Type` are mandatory for every table.
The other options allow for further customization of the tables.
For the various options how to modify the crosstabs, 
please have a look at `vignette("questions-parameters")`.


### Use the mapping

Next, we'll show how we can apply the data modifications and calculate the crosstabs in the mapping.

#### Modify data

The data is modified by applying the commands in the following sheets (using [`datadaptor::Mapping$modify_data()`](https://urswilke.gitlab.io/datadaptor/reference/Mapping.html#method-modify-data-) under the hood):

* `Variables`: This modifies the data set's variables and their labels.
* `Label`: This modifies the data set's variables' value labels and allows to generate summary variables.
* `Free1`: This allows to define additional commands modifying the data. 
All the modifications of the `Variables` & `Label` sheets could also be done there.

See [here](https://urswilke.gitlab.io/datadaptor/articles/command_blocks.html)
for the documentation of the commands used in these sheets.

#### Calculate crosstabs

The following sheets control, which and how crosstabs are calculated:

* `Macro`: This allows to define the additional header variables for the x-axis of the crosstabs.
* `Questions`: The rows in this table define the crosstabs that are calculated.

#### Execute

We can apply the data modifications and calculate the crosstabs in one command
by constructing an R6 object of class "Tabula":

```{r, message=FALSE}
m <- crosstabser::Tabula$new(
  dat = df, 
  mapping_file = mapping_file
)
```

### Alternative use without Excel

Instead of using an Excel mapping file, 
we can also define the mapping entirely in R 
without the need of an Excel file
by passing the relevant data as a list:

```{r}
mapping_file <- list(
  Variables = Variables,
  Label = Label,
  Free1 = Free1,
  Questions = Questions, 
  # You can pass the parameters of the data.frame 
  # in the `Macro` sheet as a named list:
  Macro = list(ColVar = ColVar)
)
m <- crosstabser::Tabula$new(
  dat = df, 
  mapping_file = mapping_file
)
```


